---
title: "Who is Happier?"
author: "Daniel Lam (drl2163)"
runtime: shiny
output: html_document
---

## Introduction

For this project I will utilize sentiment analysis to gain insights into the HappyDB database that helps answer the question about which groups of people are happier. To do this, I will use the sentimentr package to assign a sentiment score to each entry in the HappyDB database with a more positive score correlating with a more positive sentiment, a more negative score correlating with a more negative sentiment and a score of 0 correlating with a neutral sentiment. I will then split the data on the binary categories of marital, parenthood, and gender.
```{r load libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(sentimentr)
library(ggplot2)
library(shiny)
```

## Import data
```{r}
joined_data <- read_csv("../output/joined_data.csv")
```


## Conduct sentiment analysis using the sentimentr package
```{r}
sentiment_scores = sentiment_by(joined_data$original_hm, by = NULL)
joined_data$sent_score = sentiment_scores$ave_sentiment
```

## Examine data as a whole
```{r}
cat("Average Sentiment Score: ", mean(joined_data$sent_score), "\nMedian Sentiment Score: ", median(joined_data$sent_score))

ggplot(joined_data, aes(x = sent_score)) +    #fill = "blue", color = "black"
  geom_density(fill = "blue") +
  labs(title = "Sentiment Scores: All Data", x = "Score", y = "Density") +
  theme_minimal()
```

The average and median sentiment scores indicate a positive sentiment throughout the database. I opted to use density plots instead of histograms throughout this project as it will make the slight differences between the different group sentiment scores more noticeable. The density of the sentiment scores is bi-modal with the first maximum representing the large number of neutral entries which is explained by most of the entries describing an event/occurrence without using emotion-based words. The next maximum is in line with the average and median sentiment scores that were calculated.

## Split the data on marital status
```{r}
df_married = subset(joined_data, marital == 'married')
df_single = subset(joined_data, marital == 'single')

cat("Married entries: ", dim(df_married)[1], "\nSingle entries: ", dim(df_single)[1])
cat("Average Married Sentiment Score: ", mean(df_married$sent_score), "\nMedian Married Sentiment Score: ", median(df_married$sent_score))
cat("Average Single Sentiment Score: ", mean(df_single$sent_score), "\nMedian Single Sentiment Score: ", median(df_single$sent_score))
```
There are more single entries in than married in the database. The data indicates that married people have slightly more positive entries in the database than single people. 

## Split the data on parenthood status
```{r}
df_parent = subset(joined_data, parenthood == 'y')
df_notparent = subset(joined_data, parenthood == 'n')

cat("Parent entries: ", dim(df_parent)[1], "\nNon-parent entries: ", dim(df_notparent)[1])
cat("Average Parent Sentiment Score: ", mean(df_parent$sent_score), "\nMedian Parent Sentiment Score: ", median(df_parent$sent_score))
cat("Average Non-parent Sentiment Score: ", mean(df_notparent$sent_score), "\nMedian Non-parent Sentiment Score: ", median(df_notparent$sent_score))
```



## Split the data on gender
```{r}
df_m = subset(joined_data, gender == 'm')
df_f = subset(joined_data, gender == 'f')

cat("Average Male Sentiment Score: ", mean(df_m$sent_score), "\nMedian Male Sentiment Score: ", median(df_m$sent_score))
cat("Average Female Sentiment Score: ", mean(df_f$sent_score), "\nMedian Female Sentiment Score: ", median(df_f$sent_score))
```

## Visualization of split on binary categories
```{r}
# Define the UI
shinyApp(
ui <- fluidPage(
  titlePanel("Density Plots"),
  sidebarLayout(
    sidebarPanel(
      selectInput(
        "fill_group",
        "Choose Group:",
        choices = c("marital", "parenthood", "gender"),
        selected = "marital")
    ),
    mainPanel(
      plotOutput("densityplot")
    )
  )
),
# Define the server
server <- function(input, output) {
  # Create a reactive expression to generate the scatterplot
  densityplot_data <- reactive({
    ggplot(joined_data, aes(x = sent_score, fill = !!sym(input$fill_group))) +
      geom_density(alpha = 0.5) +
      labs(x = "Sentiment Score", y = "Frequency") +
      theme_minimal()
  })
  # Render the scatterplot
  output$densityplot <- renderPlot({
    print(densityplot_data())
  })
}
)

```


## Exploring sentiment score with age
```{r}
age_df = subset(joined_data, age <=100)
```

## Visualization of sentiment score and age
```{r}
# Define the UI
shinyApp(
ui <- fluidPage(
  titlePanel("Age and Sentiment Score"),
  sidebarLayout(
    sidebarPanel(
      selectInput(
        "color_group",
        "Choose Color Group:",
        choices = c("marital", "parenthood", "gender"),
        selected = "marital")
    ),
    mainPanel(
      plotOutput("scatterplot")
    )
  )
),
# Define the server
server <- function(input, output) {
  # Create a reactive expression to generate the scatterplot
  scatterplot_data <- reactive({
    ggplot(age_df, aes(x = age, y = sent_score, color = !!sym(input$color_group))) +
      geom_point() +
      labs(x = "Age", y = "Sentiment Score", color = input$color_group) +
      theme_minimal()
  })
  # Render the scatterplot
  output$scatterplot <- renderPlot({
    print(scatterplot_data())
  })
}
)
```












